### 2025-02-13 causalDT-manuscript/README.Rmd
add readmes

![](causalDT/man/figures/cdt_diagram.png)

@@

## Organization

This repository contains:

1. An R package `causalDT` to run causal distillation trees on your own data (see [causalDT/](causalDT/))
2. All code necessary to reproduce the analysis and figures in [Huang et al. (2025)](https://arxiv.org/abs/2502.07275) (see [causalDT-manuscript/](causalDT-manuscript/))

## Installation of the R package

You can install the `causalDT` R package via:

``` r
# install.packages("devtools")
devtools::install_github("tiffanymtang/causalDT", subdir = "causalDT")
```

## Example Usage

To illustrate an example usage of `causalDT`, we will use the AIDS Clinical Trials Group Study 175 (ACTG 175), a randomized controlled trial to determine the effectiveness of monotherapy compared to combination therapy on HIV-1-infected patients. This data can be found in the `speff2trial` R package.

```{r}
# install.packages("speff2trial")
library(speff2trial)

data <- speff2trial::ACTG175 |>
  dplyr::filter(arms %in% c(0, 2))

# pre-treatment covariates data
X <- data |> 
  dplyr::select(
    age, wtkg, hemo, homo, drugs, karnof, race, 
    gender, symptom, preanti, strat, cd80
  ) |> 
  as.matrix()
# treatment indicator variable
Z <- data |>
  dplyr::pull(treat)
# response variable
Y <- data |>
  dplyr::pull(cens)
```

Given the pre-treatment covariates data $X$, the treatment variable $Z$, and the response variable $Y$, we can run CDT as follows:

```{r fig.width=12}
library(causalDT)

set.seed(331)
causal_forest_cdt <- causalDT(
  X = X, Y = Y, Z = Z,
  teacher_model = "causal_forest"
)

plot_cdt(causal_forest_cdt)
```

Note that when using CDT, a teacher model must be chosen (the default is a causal forest). To help researchers select an appropriate teacher model, the Jaccard subgroup stability index (SSI) was introduced in [Huang et al. (2025)](https://arxiv.org/abs/2502.07275). Generally, a higher Jaccard SSI indicates a better teacher model. This teacher model selection procedure can be run as follows:

```{r}
# selecting between causal forest versus rboost
rboost_cdt <- causalDT(
  X = as.matrix(X), Y = Y, Z = Z,
  teacher_model = rboost
)
plot_jaccard(`Causal Forest` = causal_forest_cdt, `Rboost` = rboost_cdt)
```

@@

```
@article{huang2025distilling,
  title={Distilling heterogeneous treatment effects: Stable subgroup estimation in causal inference}, 
  author={Melody Huang and Tiffany M. Tang and Ana M. Kenney},
  year={2025},
  eprint={2502.07275},
  archivePrefix={arXiv},
  primaryClass={stat.ME},
  url={https://arxiv.org/abs/2502.07275}, 
}
```

>>>>

![](../causalDT/man/figures/cdt_diagram.png)

@@

## Project Structure

This directory contains all of the code necessary to reproduce the analysis and figures in [Huang et al. (2025)](https://arxiv.org/abs/2502.07275).

Moreover, to facilitate reproducibility of this work, we leveraged [`renv`](https://rstudio.github.io/renv/articles/renv.html) to create a reproducible environment and [`simChef`](https://yu-group.github.io/simChef/) to create a reproducible and transparent simulation workflow. Thus, to install the necessary dependencies and reproduce the simulation, the following steps are recommended:

1. Clone this repository.
2. Open the `causalDT-manuscript.Rproj` in RStudio.
3. Run the following code in the R console:

```r
# install.packages("renv")
renv::restore()
```

4. Run each of the simulation driver scripts in the `meals/` directory.
5. Render all of the results into a single summary `simChef` documentation by running the `meals/00_render_docs.R` script.

These steps will generate all of the figures and results from the mansucript. The rendered `simChef` documentation can be found in the `results/` directory (see [here](results/Results.html)).

@@

```r
@article{huang2025distilling,
  title={Distilling heterogeneous treatment effects: Stable subgroup estimation in causal inference}, 
  author={Melody Huang and Tiffany M. Tang and Ana M. Kenney},
  year={2025},
  eprint={2502.07275},
  archivePrefix={arXiv},
  primaryClass={stat.ME},
  url={https://arxiv.org/abs/2502.07275}, 
}
```
